A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-06-24 tex/sec4-pre-synthesis.tex
[ tex ] A bit more S4

@@

\begin{proof}
By induction on~$t$, we prove the rest of the theorem statement
\begin{equation}\label{eq:preprocess}
\forall d.~\exists v, e.~V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}
\end{equation}
in conjunction with
\begin{equation}\label{eq:preprocess'}
V |- t^{\syn\,\True\,\True} ~~\vee~~ V |- t^{\chk\,\True\,\True} ~~\vee~~ \exists e.~V |- t^{\chk\,\False\,e}
\end{equation}
Intuitively, statement~(\ref{eq:preprocess'}) gives a precise classification of a term as either a `real' synthesising term (i.e.~without missing annotations or an outermost mode cast), a `real' checking term, or a term that has missing annotations.
After a derivation is constructed for statement~(\ref{eq:preprocess'}), we can perform some slight adjustment to obtain a desired derivation for statement~(\ref{eq:preprocess}) depending on~$d$:
\begin{itemize}
\item For the first disjunct with $\dir{d} = {\syn}$, the second disjunct with $\dir{d} = {\chk}$, and the third disjunct with $\dir{d} = {\chk}$:
The mode we have matches the mode to be established, so no adjustment is needed.
\item For the first disjunct with $\dir{d} = {\chk}$:
Apply $\ChkRule{Sub}$.
\item For the second disjunct with $\dir{d} = {\syn}$:
Apply $\SynRule{Missing}$.
\item For the third disjunct with $\dir{d} = {\syn}$:
If $e = \False$, then the outermost rule of the derivation must be $\ChkRule{Sub}$, removing which we get a derivation of $V |-_{\Sigma,\Omega} t^{\syn\,\False\,\True}$;
otherwise $e = \True$, in which case we can apply $\SynRule{Missing}$.
\end{itemize}
Now we prove statement~(\ref{eq:preprocess'}) with a case analysis on~$t$.
\begin{itemize}
\item $\Rule{Var}$:
Establish the first disjunct using $\SynRule{Var}$.
\item $\Rule{Anno}$ ($t \bbcolon A$):
By statement~(\ref{eq:preprocess}) in the induction hypothesis, there is a derivation of $V |-_{\Sigma,\Omega} t^{\chk\,v\,e}$ for some boolean $v$~and~$e$.
If $v = \False$, establish the third conjunct using $\SynRule{Anno}$ and $\ChkRule{Sub}$; otherwise, establish the first conjunct using $\ChkRule{Sub}$.
\item $\Rule{Op}$:
By statement~(\ref{eq:preprocess}) in the induction hypothesis, there is a derivation of $V |-_{\Sigma,\Omega} t_i^{\dir{d_i}\,v_i\,e_i}$ for all the sub-terms~$t_i$.
\end{itemize}
\end{proof}

>>>>

\begin{lemma}\label{thm:adjustment}
For any raw term $V |-_{\Sigma,\erase\Omega} t$, if
\[ V |- t^{\syn\,\True\,\True} ~~\vee~~ V |- t^{\chk\,\True\,\True} ~~\vee~~ \exists e.~V |- t^{\chk\,\False\,e} \]
then for any mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}$ for some boolean $v$~and~$e$.
\end{lemma}

Intuitively, the antecedent gives a precise classification of a term as either a `real' synthesising term (i.e.~without missing annotations or an outermost mode cast), a `real' checking term, or a term that has missing annotations.
The lemma only performs some slight adjustment to obtain a derivation with any required mode.

\begin{proof}[Proof of \cref{thm:adjustment}]
Case analysis on the antecedent and~$\dir{d}$.
\begin{itemize}
\item For the first disjunct with $\dir{d} = {\syn}$, the second disjunct with $\dir{d} = {\chk}$, and the third disjunct with $\dir{d} = {\chk}$:
The mode we have matches the mode to be established, so no adjustment is needed.
\item For the first disjunct with $\dir{d} = {\chk}$:
Apply $\ChkRule{Sub}$.
\item For the second disjunct with $\dir{d} = {\syn}$:
Apply $\SynRule{Missing}$.
\item For the third disjunct with $\dir{d} = {\syn}$:
If $e = \False$, then the outermost rule of the derivation must be $\ChkRule{Sub}$, removing which we get a derivation of $V |-_{\Sigma,\Omega} t^{\syn\,\False\,\True}$;
otherwise $e = \True$, in which case we can apply $\SynRule{Missing}$.
\vspace{-\topsep-\baselineskip}
\end{itemize}
\end{proof}

@@

\begin{proof}
The conclusion is the same as that of \cref{thm:adjustment}, so it suffices to prove the antecedent of \cref{thm:adjustment} by induction on~$t$.
\begin{itemize}
\item $\Rule{Var}$:
Establish the first disjunct using $\SynRule{Var}$.
\item $\Rule{Anno}$ ($t \bbcolon A$):
By the induction hypothesis and \cref{thm:adjustment}, there is a derivation of $V |-_{\Sigma,\Omega} t^{\chk\,v\,e}$ for some boolean $v$~and~$e$.
If $v = \False$, establish the third conjunct using $\SynRule{Anno}$ and $\ChkRule{Sub}$; otherwise, establish the first conjunct using $\ChkRule{Sub}$.
\item $\Rule{Op}$ ($\tmOpts$):
By the induction hypotheses and \cref{thm:adjustment}, there is a derivation of $V, \vec x_i |-_{\Sigma,\Omega} t_i^{\dir{d_i}\,v_i\,e_i}$ for each sub-term~$t_i$, and by applying $\Rule{Op}$ we get a derivation of $V |-_{\Sigma,\Omega} \tmOpts^{\dir d\,v\,\True}$ where $v = \bigwedge_i v_i$.
If $v = \True$, establish the first or second disjunct depending on~$\dir d$; otherwise, establish the third disjunct, applying $\ChkRule{Sub}$ if $\dir d = {\syn}$.
\vspace{-\topsep-\baselineskip}
\end{itemize}
\end{proof}
